---
title: "Foraging Activity Budget Scan Analysis 2007-2018"
output: html_document
---


```

## R Markdown 


---
title: "Foraging Activity Budget Script for Combined Scan Data 2007-2018"
output: html_notebook

#missing dom data for LV (2007), PR (2016), and SF (2017)

#removed Helen from dominance and reproductive status since this dataset is subsetted by adult females

#behavioral data file has PR removed for 2016 (no dominance data) and SF removed for 2017 (disappeared mid season). LV (Lavender, not the group named LV) rank has been assumed same as 2007 for 2008

#removed erroneous scans where individual from a different group was mistakenly recorded (ie, observing Kiara when cleary the researcher was with LV)

#rotations split when RS changes. This occurred 8 times. For one AD rotation, 2 monkeys changed RS and rotation was split into 3. Rotation also split during extreme weather events (>10 cm of rain in 24 hrs)
---
```{r}
memory.limit(size=9000)

library(dplyr)
library(ggplot2)
library(lme4)
library(emmeans)
library(lubridate)
library(tidyverse)
library(fuzzyjoin)
library(car)
library(xlsx)
library(gridExtra)
library(grid)
library(gtable)
library(ggpubr)
library(DataCombine)
library(see)
library(extrafont)
library(multcompView)
library(ggpubr)
library(lsmeans)
library(multcomp)
library(tidytext)

```
#budget analysis

```{r}
#reading in reproductive status (RS) table that contains RS data for every adult female from the years 2007, 2008, 2016, 2017, 2018
repStatusTable <- read.csv("RepStatusTable_withoutHL_UpdatedMay4.csv") 

#renaming MonkeyName column
repStatusTable <- repStatusTable %>% 
           rename(
           MonkeyName = ï..MonkeyName,
   
                         )
#setting as DF object
repStatusTable <- as.data.frame(repStatusTable)

#Setting StartDate and EndDate as Date objects to facilitate join
repStatusTable$StartDate <- parse_date_time(repStatusTable$StartDate, orders = "mdy")
repStatusTable <- repStatusTable %>% mutate(StartDate = as_datetime(StartDate))
repStatusTable$StartDate <- as.Date(repStatusTable$StartDate)
repStatusTable$EndDate <- parse_date_time(repStatusTable$EndDate, orders = "mdy")
repStatusTable <- repStatusTable %>% mutate(EndDate = as_datetime(EndDate))
repStatusTable$EndDate <- as.Date(repStatusTable$EndDate)


#Reading in Dominance table that contains rankings for every adult female for the years 2007, 2008, 2016, 2017, 2018
dominanceTable <- read.csv("Dom_20072018.csv")
dominanceTable <- as.data.frame(dominanceTable)

#Setting DateStart and DateEnd as Date objects to facilitate the join
dominanceTable$DateStart <- parse_date_time(dominanceTable$DateStart, orders = "mdy")
dominanceTable <- dominanceTable %>% mutate(DateStart = as_datetime(DateStart))
dominanceTable$DateStart <- as.Date(dominanceTable$DateStart)
dominanceTable$DateEnd <- parse_date_time(dominanceTable$DateEnd, orders = "mdy")
dominanceTable <- dominanceTable %>% mutate(DateEnd = as_datetime(DateEnd))
dominanceTable$DateEnd <- as.Date(dominanceTable$DateEnd)

#Reading in behavioral data that contains 62,251 scans 
behavData <- read.csv("Scans_Combined_Mar10.csv") 
behavData <- as.data.frame(behavData)

#Removing duplicate monkeys for each unique date + time combo. Now 61,569 scans
behavData <- behavData %>% distinct(MonkeyCode, DateTime, .keep_all = TRUE)

#Subsetting to just adult females, which whittles it down to 22,053 scans
behavData<- behavData[behavData$Sex == "F" & behavData$Age == "A",]


#Removing unused levels
behavData <- droplevels(behavData)

#Setting DateTime as a DateTime object
behavData$DateTime <- parse_date_time(behavData$DateTime, orders = "mdy HM")
behavData <- behavData %>% mutate(DateTime = as_datetime(DateTime))
behavData$DateTime <- as.Date(behavData$DateTime)


#Reading in .csv that contains rotation numbers for each date of observation
combinedDF_rot <- read.csv("Rotation_Dates_UpdatedMay13.csv")
combinedDF_rot <- as.data.frame(combinedDF_rot)

#Setting DateTime as a date object
combinedDF_rot$DateOfYear <- parse_date_time(combinedDF_rot$DateOfYear, orders = "mdy")
combinedDF_rot <- combinedDF_rot %>% mutate(DateTime = as_datetime(DateOfYear)) %>%
  dplyr::select(-DateOfYear)
combinedDF_rot$DateTime <- as.Date(combinedDF_rot$DateTime)


combinedDF_rot <- combinedDF_rot %>%

                 dplyr::select(-X) %>% 
            
                 distinct()

#Joining rotation numbers to behavioral data
behavData <- behavData %>% left_join(., combinedDF_rot, by = c("DateTime" = "DateTime"))
                                                            
#Setting rotation as a factor
behavData$Rotation <- as.factor(behavData$Rotation)
behavData$Phenotype <- as.factor(behavData$Phenotype)
behavData$MonkeyCode <- as.factor(behavData$MonkeyCode)
behavData$Group <- as.factor(behavData$Group)
#Fuzzy left joining reproductive status to behavioral data
behavDataRepStatus <- behavData %>%
            fuzzy_left_join(., repStatusTable, by = c("MonkeyCode" = "MonkeyCode", "DateTime" = "StartDate", "DateTime" = "EndDate"), match_fun = list(`==`, `>=`, `<=`))

#Removing extra columns that fuzzy join adds
behavDataRepStatus <- behavDataRepStatus %>%
                      dplyr::select(-MonkeyName.y, -MonkeyCode.y, -StartDate, -EndDate) 


#Renaming MonkeyCode
behavDataRepStatus <- behavDataRepStatus %>%
              rename(MonkeyCode = MonkeyCode.x)

#Renaming Monkey Name
behavDataRepStatus <- behavDataRepStatus %>%
              rename(MonkeyName = MonkeyName.x)

#setting Rep Status as factor
behavDataRepStatus$ReproductiveStatus <- as.factor(behavDataRepStatus$ReproductiveStatus)

#Fuzzy left joining dominance to new behavioral data DF
combinedDF <- behavDataRepStatus %>% 
              fuzzy_left_join(., dominanceTable, 
                              by = c("MonkeyCode" = "MonkeyCode", "DateTime" = "DateStart", "DateTime" = "DateEnd"), match_fun = 
                              list(`==`, `>=`, `<=`))




#Removing columns we don't need, and renaming "MonkeyCode.x"
combinedDF <- combinedDF %>%
              dplyr::select(-MonkeyCode.y, -DateStart, -DateEnd)
combinedDF <- combinedDF %>%
              rename(MonkeyCode = MonkeyCode.x)


#Reading in DF that contains temperature and rainfall values for 2007, 2008, 2016, 2017, 2018 study periods
weatherData <- read.csv("WeatherData_2007_2018.csv") 
weatherData <- as.data.frame(weatherData)
weatherData$DateTime <- parse_date_time(weatherData$DateTime, orders = "mdy")
weatherData <- weatherData %>% mutate(DateTime = as_datetime(DateTime))
weatherData$DateTime <- as.Date(weatherData$DateTime)

#Joining weather data to combinedDF
combinedDF1 <- combinedDF %>% left_join(., weatherData, by = c("DateTime" = "DateTime"))

#Grouping by variables of interest
df_budget <- combinedDF1 %>% 
  group_by(DateTime, MonkeyCode, MonkeyName, Phenotype, Group, State, DominanceScale, ReproductiveStatus, TempMax, TempMin, Rain) 
  

#These lines count each state for each monkey per rotation and spread them so that one line = one monkeys combined states per rotation
df_budget <- df_budget %>%  summarise(Count=n())
df_budget <- df_budget %>% ungroup()
df_budget <- df_budget %>% spread(State,Count)

#Joining rotation to df_budget again since it was lost with group by, setting rotation as factor
df_budget <- df_budget %>% left_join(., combinedDF_rot, by = c("DateTime" = "DateTime"))
df_budget$Rotation <- as.factor(df_budget$Rotation)

#Setting NA's to zero
df_budget[c("EFI", "FFR", "FIN", "FOT","RES", "SAC","TRA","VFO", "OTH", "SDI", "PLA", "SAG", "VIG", "DRI")][is.na(df_budget[c("EFI", "FFR", "FIN", "FOT","RES", "SAC", "TRA","VFO", "OTH", "SDI", "PLA", "SAG", "VIG", "DRI")])] <- 0
```


```{r}
#Z-transforming ecological variables
df_transformed <- df_budget %>%
       mutate_at(vars("TempMax", "TempMin", "Rain"), list(~as.numeric(scale(.)))) 

#grouping by rotation and monkey code, summing scans in each state, and averaging weather variables over the rotation, resulting in one row per monkey per rotation
df_budget2 <- df_transformed %>%
              group_by(Rotation, MonkeyCode) %>%
              summarise(
                        Group = unique(Group),
                        TempMax = mean(TempMax),
                        TempMin = mean(TempMin),
                        Rain = mean(Rain),
                        DRI = sum(DRI),
                        EFI = sum(EFI),
                        FFR = sum(FFR),
                        FIN = sum(FIN),
                        FOT = sum(FOT),
                        OTH = sum(OTH),
                        PLA = sum(PLA),
                        RES = sum(RES),
                        SAC = sum(SAC),
                        SAG = sum(SAG),
                        SDI = sum(SDI),
                        TRA = sum(TRA),
                        VFO = sum(VFO),
                        VIG = sum(VIG),
                        DominanceScale = mean(DominanceScale),
                        ReproductiveStatus = unique(ReproductiveStatus),
                        Phenotype = unique(Phenotype))

                         #if rotations spanned the end of a month, the month the rotation started is included in the grouped df



#Creating a new ordinal variable, DominanceCategory, from DominanceScale
df_budget2$DominanceCategory <- if_else(df_budget2$DominanceScale <= 0.33, "low", if_else(df_budget2$DominanceScale > 0.33 & df_budget2$DominanceScale <= 0.66, "mid","high"))
df_budget2$DominanceCategory <- as.factor(df_budget2$DominanceCategory)

#This line takes the original behavioral dataset, groups by monkey and rotation, and counts how many scans each monkey was observed per rotation
totals <- behavData %>% group_by(Rotation = as.factor(Rotation), MonkeyCode = as.factor(MonkeyCode)) %>% 
  summarise(TotalScans=n())

#Joining TotalScans  column from above to df_budget, so that every monkey per rotation has their total scans from that rotation in a seperate column
df_budget2 <- df_budget2 %>% left_join(., totals, by = c("Rotation" = "Rotation", 
                                                            "MonkeyCode" = "MonkeyCode"))


#Filtering only monkeys that have 3 or greater total scans per rotation
df_budget2 <- df_budget2[!(df_budget2$TotalScans < 3),]



#Creating new column that is total foraging scans
df_budget2$TotalForaging <- c(df_budget2$EFI + df_budget2$FFR +
                                          df_budget2$FIN + df_budget2$FOT)
                                          

#Replacing "D" and "T" with "Dichromat" and "Trichromat"
df_budget2$Phenotype <- Recode(df_budget2$Phenotype, "'D'='Dichromat'; 'T'='Trichromat'")
```

#General foraging model, looking at total foraging scans (as proportion of total scans) in response to phenotype, dominance, RS, and controlling for monkey ID and group

```{r}

GeneralForagingModel <- glmer(TotalForaging ~  
                                Phenotype + 
                                DominanceCategory +
                                ReproductiveStatus +
                                TempMax +
                                TempMin +
                                Rain +
                                offset(log(TotalScans)) + 
                                (1 | Group/MonkeyCode),
                              data = df_budget2,
                              family = poisson(link = "log"))

summary(GeneralForagingModel)
drop1(GeneralForagingModel, test = "Chisq")
Anova(GeneralForagingModel)


#Statistics on the model
ForagingModel.lsm <- emmeans(GeneralForagingModel, "Phenotype", type = "response", adjust = "tukey") #This function displays the LSM

pairs(ForagingModel.lsm)
pairs(emmeans(GeneralForagingModel, "Phenotype", adjust = "tukey"))

ForagingModel.cld <- cld(ForagingModel.lsm)

ForagingModel.lsm.cld.tukey <- cld.emmGrid(ForagingModel.lsm, alpha = 0.05, Letters = letters, adjust = "tukey") #This function creates a compact letter display of the LSM enabling use with ggplot. It also adds a Tukey adjustment to test for honest significant difference between the means.
   
```


#Surface-dwelling insect foraging model, looking at insect foraging scans (as proportion of total scans) in response to phenotype, dominance, RS, and controlling for monkey ID and group
``` {r}

InsectForagingModel <- glmer(FIN ~  
                               
                                Phenotype + 
                                DominanceCategory +
                               ReproductiveStatus +
                               TempMax +
                               TempMin + 
                               Rain +
                                offset(log(TotalScans)) + 
                                (1 | Group/MonkeyCode), 
                              data = df_budget2,
                              family = poisson(link = "log"),control = glmerControl(optCtrl = list(maxfun = 1500000)))


summary(InsectForagingModel)
drop1(InsectForagingModel, test = "Chisq")
Anova(InsectForagingModel)

#Statistics on the model
InsectModel.lsm <- emmeans(InsectForagingModel, "Phenotype", type = "response", adjust = "tukey") #This function displays the LSM   

pairs(InsectModel.lsm)
pairs(emmeans(InsectForagingModel, "Phenotype", adjust = "tukey"))

```


#Fruit foraging model, looking at fruit foraging scans (as proportion of total scans) in response to phenotype, dominance, RS, and controlling for monkey ID and group
```{r}
FruitFeedingModel <- glmer(FFR ~  
                                 
                                   Phenotype + 
                                   DominanceCategory +
                                    ReproductiveStatus +
                                    TempMax + 
                                    TempMin +
                                    Rain +
                                   offset(log(TotalScans)) + 
                                   (1 | Group/MonkeyCode), 
                                 data = df_budget2,
                                 family = poisson(link = "log"),control = glmerControl(optCtrl = list(maxfun = 1500000)))
summary(FruitFeedingModel)
drop1(FruitFeedingModel, test = "Chisq")
Anova(FruitFeedingModel)


#Statistics on the model

FruitModel.lsm <- emmeans(FruitFeedingModel, "Phenotype", type = "response", adjust = "tukey") #This function displays the LSM

pairs(FruitModel.lsm)
pairs(emmeans(FruitFeedingModel, "Phenotype", adjust = "tukey"))

```


###The chunk below takes the finalized, but ungrouped, dataframe (with all relevant variables added) and adds fruit taxon color data to analyze scans spent on specifically on conpsicuous fruits (to TRICHROMATS only!)
```{r}
fruit_cat_df <- combinedDF1

fruit_cat_df <- fruit_cat_df %>%
  dplyr::select(-ColourCategory) 

#Subset of scan data that is foraging
df_forage <- fruit_cat_df[fruit_cat_df$State == "VFO"|fruit_cat_df$State == "FFR"|fruit_cat_df$State == "EFI"|fruit_cat_df$State == "FIN"|fruit_cat_df$State =="FOT",]

#Subset of scan data that is fruit feeding
df_fruit <- df_forage[df_forage$State == "FFR",]

#Removing unused levels
df_fruit <- droplevels(df_fruit)

df_color <- read.csv("colors_final_final.csv")
df_fruit <- df_fruit %>% left_join(., df_color, by = c("Taxon" = "Taxon" 
))

#subsetting by only fruits with assigned color categories
df_fruit <- df_fruit[df_fruit$ColourCategory == "con" |df_fruit$ColourCategory == "cry" | df_fruit$ColourCategory == "dark" | df_fruit$ColourCategory == "con_both"  ,]


#Groups the data based on DateTime, MonkeyCode, MonkeyName, Phenotype, Group, ColourCategory, DominanceScale, ReproductiveStatus, TempMax, TempMin, and Rain

df_fruit_grouped <- df_fruit %>% 
  group_by(DateTime, MonkeyCode, MonkeyName, Phenotype, Group, ColourCategory, DominanceScale, ReproductiveStatus, TempMax, TempMin, Rain) 


#These lines count each state for each monkey per rotation and spread them so that one line = one monkeys combined states per DateTime
df_fruit_grouped <- df_fruit_grouped %>%  summarise(Count=n())
df_fruit_grouped <- df_fruit_grouped %>% ungroup()
df_fruit_grouped <- df_fruit_grouped %>% spread(ColourCategory,Count)



#Joining the rotation DF 
df_fruit_grouped <- df_fruit_grouped %>% left_join(., combinedDF_rot, by = c("DateTime" = "DateTime"))

#Setting NA's to zero
df_fruit_grouped[c("con","con_both", "cry", "dark")][is.na(df_fruit_grouped[c("con", "con_both", "cry", "dark")])] <- 0



#Transforming ecological variables
df_fruit_transformed <- df_fruit_grouped %>%
  mutate_at(vars("TempMax", "TempMin", "Rain"), list(~as.numeric(scale(.)))) 


#grouping by rotation and monkey code, summing fruit conspicuity scans, and averaging weather variables over the rotation, resulting in one row per monkey per rotation
df_cat <- df_fruit_transformed %>%
  
  group_by(Rotation, MonkeyCode) %>%
  
  summarise(
    
    Group = unique(Group),
    
    TempMax = mean(TempMax),
    
    TempMin = mean(TempMin),
    
    Rain = mean(Rain),
    
    con = sum(con),
    
    con_both = sum(con_both),
    
    cry = sum(cry),
    
    dark = sum(dark),
    
    DominanceScale = mean(DominanceScale),
    
    ReproductiveStatus = unique(ReproductiveStatus),
    
    Phenotype = unique(Phenotype))




#The next two lines create total scans column that shows you how many scans that individual was observed in each day
totals_cat <- fruit_cat_df %>% group_by(Rotation, MonkeyCode = as.factor(MonkeyCode)) %>% 
summarise(TotalScans=n())

df_cat_final <- df_cat %>% left_join(., totals_cat, by = c("Rotation" = "Rotation", 
                                                                "MonkeyCode" = "MonkeyCode"
                                                                ))
#Subsetting to include monkeys that have 3 or greater total scans per rotation
df_cat_final <- df_cat_final[!(df_cat_final$TotalScans < 3) ,]

#Recoding the phenotype variable 
df_cat_final$Phenotype <- Recode(df_cat_final$Phenotype, "'D'='Dichromat'; 'T'='Trichromat'")

#Creating a new ordinal variable, DominanceCategory, from DominanceScale

df_cat_final$DominanceCategory <- if_else(df_cat_final$DominanceScale <= 0.33, "low", if_else(df_cat_final$DominanceScale > 0.33 & df_cat_final$DominanceScale <= 0.66, "mid","high"))

#Adding a column for scans for fruit that are equally conspicuous or inconspicuous to dichromats and trichromats
df_cat_final <- df_cat_final %>% mutate(equal_discrimination = dark + cry + con_both)

#Removing a row that contains only NAs
df_cat_final <- na.omit(df_cat_final)

#Modeling number of scans spent conspicuous fruit foraging

GeneralConFruitModel <- glmer(con ~  
                                ReproductiveStatus +
                                DominanceCategory +
                                Phenotype + 
                                TempMax +
                                TempMin + 
                                Rain +
                                offset(log(TotalScans)) + 
                                (1 | Group/MonkeyCode), 
                              data = df_cat_final,
                              family = poisson(link = "log"), control = glmerControl(optCtrl = list(maxfun = 1500000)) )

summary(GeneralConFruitModel)
drop1(GeneralConFruitModel, test = "Chisq")
Anova(GeneralConFruitModel)

#Statistics on the model

ConModel.lsm <- emmeans(GeneralConFruitModel, "Phenotype", type = "response", adjust = "tukey") #This function displays the LSM

pairs(emmeans(GeneralConFruitModel, "Phenotype", adjust = "tukey"))


#Rerunning conspicuous fruit model with an interaction between phenotype and dominance
Model_interaction <- glmer(con ~  
                                ReproductiveStatus +
                                DominanceCategory:Phenotype +
                                DominanceCategory +
                                Phenotype +
                                TempMax +
                                TempMin + 
                                Rain +
                                offset(log(TotalScans)) + 
                                (1 | Group/MonkeyCode), 
                              data = df_cat_final,
                              family = poisson(link = "log"), control = glmerControl(optCtrl = list(maxfun = 1500000)) )

summary(Model_interaction)
drop1(Model_interaction, test = "Chisq")

```

